\documentclass[11pt, a4paper, titlepage]{article}
\usepackage{amsmath}
\usepackage{a4wide}
\usepackage{url}
\usepackage{multirow}
\usepackage{amsfonts}
\newcommand{\release}{4.1.0}
\newcommand{\JAGS}{\textsf{JAGS}}
\newcommand{\BUGS}{\textsf{BUGS}}
\newcommand{\WinBUGS}{\textsf{WinBUGS}}
\newcommand{\R}{\textsf{R}}
\newcommand{\CODA}{\textsf{coda}}

\usepackage{verbatim}

\newcommand{\code}[1]{{\bgroup{\normalfont\ttfamily #1}\egroup}}
\newcommand{\samp}[1]{{`\bgroup\normalfont\texttt{#1}'\egroup}}
\newcommand{\file}[1]{{`\normalfont\textsf{#1}'}}
\let\command=\code
\let\option=\samp

\begin{document}

\title{JAGS Version \release\ installation manual}
\author{Martyn Plummer \and Bill Northcott \and Matt Denwood}
\date{19 January 2016}
\maketitle

\JAGS\ is distributed in binary format for Microsoft Windows, Mac OS
X, and most Linux distributions.  The following instructions are for
those who wish to build \JAGS\ from source. The manual is divided
into three sections with instructions for Linux/Unix, Mac OS X, and Windows.

\section{Linux and UNIX}

\JAGS\ follows the usual GNU convention of 
\begin{verbatim}
./configure
make
make install
\end{verbatim}
which is described in more detail in the file \texttt{INSTALL} in
the top-level source directory. On some UNIX platforms, you may
be required to use GNU make (gmake) instead of the native make
command. On systems with multiple processors, you may use the option 
\option{-j} to speed up compilation, {\em e.g.} for a quad-core PC you
may use:
\begin{verbatim}
make -j4
\end{verbatim}
If you have the cppunit library installed then you can test the build with
\begin{verbatim}
make check
\end{verbatim}
WARNING. If you already have a copy of the jags library installed on
your system then the test program created by \code{make check} will
link against the {\bf installed} library and not the one in your build
directory. So if the test suite includes a regression test for a bug
that was fixed in the version you are building but a previous version
of JAGS is already installed then the unit tests will fail. Best
practice is to run the tests after \code{make install} so the build
and installed versions are the same.

\subsection{Configure options}

At configure time you also have the option of defining options such
as:
\begin{itemize}
\item The names of the C, C++, and Fortran compilers.  
\item Optimization flags for the compilers.  \JAGS\ is optimized by
  default if the GNU compiler (gcc) is used. If you are using another
  compiler then you may need to explicitly supply optimization flags.
\item Installation directories. \JAGS\ conforms to the GNU standards
  for where files are installed. You can control the installation
  directories in more detail using the flags that are listed when
  you type \command{./configure --help}.
\end{itemize}

\subsubsection{Configuration for a 64-bit build}

By default, JAGS will install all libraries into
\file{/usr/local/lib}.  If you are building a 64-bit version of \JAGS,
this may not be appropriate for your system. On Fedora and other
RPM-based distributions, for example, 64-bit libraries should be
installed in \file{lib64}, and on Solaris, 64-bit libraries are in a
subdirectory of \file{lib} ({\em e.g.} \file{lib/amd64} if you are
using a x86-64 processor), whereas on Debian, and other Linux
distributions that conform to the FHS, the correct installation
directory is \file{lib}.

To ensure that \JAGS\ libraries are installed in the correct
directory, you should supply the \option{--libdir} argument to the
configure script, {\em e.g.}:
\begin{verbatim}
./configure --libdir=/usr/local/lib64
\end{verbatim}

It is important to get the installation directory right when using the
\code{rjags} interface between R and \JAGS, otherwise the
\code{rjags} package will not be able to find the \JAGS\ library.

\subsubsection{Configuration for a private installation}

If you do not have administrative privileges, you may wish to install
\JAGS\ in your home directory. This can be done with the following
configuration options
\begin{verbatim}
export JAGS_HOME=$HOME/jags #or wherever you want it
./configure --prefix=$JAGS_HOME
\end{verbatim}
For more detailed control over the installation directories type
\begin{verbatim}
./configure --help
\end{verbatim}
and read the section ``Fine-tuning of the installation directories.''

With a private installation, you need to modify your PATH environment
variable to include \file{\$JAGS\_HOME/bin}. You may also need to set
\code{LD\_LIBRARY\_PATH} to include \file{\$JAGS\_HOME/lib} (On Linux this
is not necessary as the location of libjags and libjrmath is hard-coded
into the \JAGS\ binary).

\subsection{BLAS and LAPACK}
\label{section:blas:lapack}

BLAS (Basic Linear Algebra System) and LAPACK (Linear Algebra Pack)
are two libraries of routines for linear algebra. They are used by the
multivariate functions and distributions in the \texttt{bugs} module.
Most unix-like operating system vendors supply shared libraries that
provide the BLAS and LAPACK functions, although the libraries may not
literally be called ``blas'' and ``lapack''.  During configuration, a
default list of these libraries will be checked. If \texttt{configure}
cannot find a suitable library, it will stop with an error message.

You may use alternative BLAS and LAPACK libraries using the configure
options \texttt{--with-blas} and \texttt{--with-lapack}
\begin{verbatim}
./configure --with-blas="-lmyblas" --with-lapack="-lmylapack"
\end{verbatim}

If the BLAS and LAPACK libraries are in a directory that is not on the
default linker path, you must set the \code{LDFLAGS} environment variable
to point to this directory at configure time:
\begin{verbatim}
LDFLAGS="-L/path/to/my/libs" ./configure ...
\end{verbatim}

At runtime, if you have linked \JAGS\ against BLAS or LAPACK in
a non-standard location, you must supply this location with the
environment variable \code{LD\_LIBRARY\_PATH}, {\em e.g.}
\begin{verbatim}
LD_LIBRARY_PATH="/path/to/my/libs:${LD_LIBRARY_PATH}"
\end{verbatim}
Alternatively, you may hard-code the paths to the blas and lapack
libraries at compile time. This is compiler and platform-specific,
but is typically achieved with
\begin{verbatim}
LDFLAGS="-L/path/to/my/libs -R/path/to/my/libs
\end{verbatim}

JAGS can also be linked to static BLAS and LAPACK if they have both
been compiled with the \texttt{-fPIC} flag. You will probably need to
do a custom build of BLAS and LAPACK if you require this. The
configure options for JAGS are then:
\begin{verbatim}
./configure --with-blas="-L/path/to/my/libs -lmyblas -lgfortran -lquadmath" \
            --with-lapack="-L/path/to/my/libs -lmylapack"
\end{verbatim}
Note that with static linking you must add the dependencies of the BLAS
library manually. The LAPACK library will pick up the same dependencies.
Note also that libtool does not like linking directly to archive files.
If you attempt a configuration of the form
\begin{verbatim}
--with-blas="/path/to/my/libs/myblas.a"
\end{verbatim}
then this will pass at configure time but ``make'' will not correctly
build the JAGS modules.

\subsubsection{Multithreaded BLAS and LAPACK}
\label{section:blas:multithreaded}

Some high-performance computing libraries offer multi-threaded
versions of the BLAS and LAPACK libraries. Although instructions for
linking against some of these libraries are given below, this should
not be taken as encouragement to use multithreaded BLAS.  Testing
shows that using multiple threads in BLAS can lead to significantly
{\em worse} performance while using up substantially more computing
resources.

\subsection{GNU/Linux}
\label{section:gnulinux}

GNU/Linux is the development platform for \JAGS, and a variety of
different build options have been explored, including the use of
third-party compilers and linear algebra libraries.

\subsubsection{Fortran compiler}

The GNU FORTRAN compiler changed between gcc 3.x and gcc 4.x from
\code{g77} to \code{gfortran}. Code produced by the two compilers is
binary incompatible. If your BLAS and LAPACK libraries are linked
against \code{libgfortran}, then they were built with \code{gfortran}
and you must also use this to compile \JAGS. 

Most recent GNU/Linux distributions have moved completely to gcc 4.x.
However, some older systems may have both compilers installed.
Unfortunately, if \code{g77} is on your path then the configure script
will find it first, and will attempt to use it to build \JAGS. This
results in a failure to recognize the installed BLAS and LAPACK
libraries. In this event, set the \code{F77} variable at configure time.
\begin{verbatim}
F77=gfortran ./configure
\end{verbatim}

\subsubsection{BLAS and LAPACK}

The {\bf BLAS} and {\bf LAPACK} libraries from Netlib
(\url{http://www.netlib.org}) should be provided as part of your Linux
distribution. If your Linux distribution splits packages into ``user''
and ``developer'' versions, then you must install the developer
package ({\em e.g.}  \texttt{blas-devel} and \texttt{lapack-devel}).

On {\bf Red Hat Enterprise Linux (RHEL)} you need to activate an
optional repository in order to have access to BLAS and LAPACK.  The
repository is called is called \texttt{rhel-<v>-<type>-optional-rpms},
where \texttt{<v>} is the RHEL release version and \texttt{<type>} is
the type of installation (server or workstation). Find the
corresponding entry in the file \file{/etc/yum.repos.d/redhat.repo}
and change the line \texttt{enabled = 0} to \texttt{enabled = 1}.

{\bf Suse Linux Enterprise Server (SLES)} does not include BLAS and
LAPACK in the main distribution. They are included in the SLES SDK, on
a set of CD/DVD images which can be downloaded from
\url{https://download.suse.com/index.jsp} (subscription and login
required).

\subsubsection{ATLAS}

On Fedora Linux, pre-compiled atlas libraries are available via the
\texttt{atlas} and \texttt{atlas-devel} RPMs.  These RPMs install the
atlas libraries in the non-standard directory \texttt{/usr/lib/atlas}
(or \texttt{/usr/lib64/atlas} for 64-bit builds) to avoid conflicts
with the standard \texttt{blas} and \texttt{lapack} RPMs. To use the
atlas libraries, you must supply their location using the
\code{LDFLAGS} variable (see section \ref{section:blas:lapack})
\begin{verbatim}
./configure LDFLAGS="-L/usr/lib/atlas"
\end{verbatim}
Runtime linking to the correct libraries is ensured by the automatic
addition of \texttt{/usr/lib/atlas} to the linker path (see the directory
\texttt{/etc/ld.so.conf.d}), so you do not need to set the
environment variable \code{LD\_LIBRARY\_PATH} at run time.

\subsubsection{AMD Core Math Library}
\label{section:acml:linux}

The AMD Core Math Library (acml) provides optimized BLAS and LAPACK
routines for AMD processors. To link \JAGS\ with \texttt{acml}, you must
supply the \texttt{acml} library as the argument to \texttt{--with-blas}.
It is not necessary to set the \texttt{--with-lapack} argument
as \texttt{acml} provides both sets of functions. See also
section~\ref{section:blas:lapack} for run-time instructions.

For example, to link to the 64-bit acml using gcc 4.0+:
\begin{verbatim}
LDFLAGS="-L/opt/acml4.3.0/gfortran64/lib" \
./configure --with-blas="-lacml -lacml_mv" 
\end{verbatim}
The \code{acmv\_mv} library is a vectorized math library that
exists only for the 64-bit version and is omitted when linking against
32-bit acml.

On multi-core systems, you may wish to use the threaded acml library
(See the warning in section \ref{section:blas:multithreaded} however).
To do this, link to \code{acml\_mp} and add the compiler flag
\option{-fopenmp}:
\begin{verbatim}
LDFLAGS="-L/opt/acml4.3.0/gfortran64_mp/lib" \
CXXFLAGS="-O2 -g -fopenmp" ./configure --with-blas="-lacml_mp -lacml_mv" 
\end{verbatim}
The number of threads used by multi-threaded acml may be controlled
with the environment variable \code{OMP\_NUM\_THREADS}.

\subsubsection{Intel Math Kernel Library}

The Intel Math Kernel library (MKL) provides optimized BLAS and LAPACK
routines for Intel processors.  MKL is designed to be linked to
executables, not shared libraries. This means that it can only be
linked to a static version of \JAGS, in which the \JAGS\ library and
modules are linked into the main executable. To build a static version
of \JAGS, use the configure option \option{--disable-shared}.

MKL version 10.0 and above uses a ``pure layered'' model for linking.
The layered model gives the user fine-grained control over four
different library layers: interface, threading, computation, and
run-time. Some examples of linking to MKL using this layered model are
given below. These examples are for GCC compilers on
\code{x86\_64}. The choice of interface layer is important on
\code{x86\_64} since the Intel Fortran compiler returns complex values
differently from the GNU Fortran compiler. You must therefore use the
interface layer that matches your compiler (\code{mkl\_intel*} or
\code{mkl\_gf*}).

For further guidance, consult the MKL Link Line advisor at
\url{http://software.intel.com/en-us/articles/intel-mkl-link-line-advisor}.

Recent versions of MKL include a shell script that sets up the
environment variables necessary to build an application with MKL.
\begin{verbatim}
source /opt/intel/composerxe-2011/mkl/bin/mklvars.sh intel64
\end{verbatim}

After calling this script, you can link \JAGS\ with a sequential
version of MKL as follows:
\begin{verbatim}
./configure --disable-shared \
  --with-blas="-lmkl_gf_lp64 -lmkl_sequential -lmkl_core -lpthread"
\end{verbatim}
Note that \code{libpthread} is still required, even when linking
to sequential MKL.

Threaded MKL may be used with:
\begin{verbatim}
./configure --disable-shared \
  --with-blas="-lmkl_gf_lp64 -lmkl_gnu_thread -lmkl_core -liomp5 -lpthread"
\end{verbatim}
The default number of threads will be chosen by the OpenMP software,
but can be controlled by setting \code{OMP\_NUM\_THREADS} or
\code{MKL\_NUM\_THREADS}.  (See the warning in section
\ref{section:blas:multithreaded} however).

\subsubsection{Using Intel Compilers}

\JAGS\ has been successfully built with the Intel Composer XE
compilers. To set up the environment for using these compilers call
the \file{compilervars.sh} shell script, {\em e.g.}
\begin{verbatim}
source /opt/intel/composerxe-2011/bin/compilervars.sh intel64
\end{verbatim}
Then call the configure script with the Intel compilers:
\begin{verbatim}
CC=icc CXX=icpc F77=ifort ./configure 
\end{verbatim}

\subsubsection{Using Clang}

\JAGS\ has been built with the clang compiler for C and C++ (version 3.1).
The configuration was
\begin{verbatim}
LD="llvm-ld" CC="clang" CXX="clang++" ./configure
\end{verbatim}
In this configuration, the gfortran compiler was used for Fortran and
the C++ code was linked to the GNU standard C++ library (libstdc++)
rather than the version supplied by the LLVM project (libc++).

\subsection{Solaris}

\JAGS\ has been successfully built and tested on the Intel x86
platform under Solaris 11.3 using the Sun Studio 12.3 and 12.4
compilers.
\begin{verbatim}
./configure CC=cc CXX=CC F77=f95 \
CFLAGS="-xO3 -xarch=sse2" \
FFLAGS="-xO3 -xarch=sse2" \
CXXFLAGS="-xO3 -xarch=sse2 -library=stlport4"
\end{verbatim}
The Sun Studio compiler is not optimized by default. Use the option
\option{-xO3} for optimization (NB This is the letter ``O'' not the
number 0) In order to use the optimization flag \option{-xO3} you
must specify the architecture with the \option{-xarch} flag. The options
above are for an Intel processor with SSE2 instructions. This must be
adapted to your own platform.

Starting with release 4.1.0, the default C++ library \file{libCstd} is
no longer supported. You must therefore add the option
\option{-library=stlport4} to \option{CXXFLAGS} to use the alternative
STLPort4 library.

%C++11?

To compile a 64-bit version of JAGS, add the option \option{-m64} to
all the compiler flags. On Solaris, 64-bit files are generally
installed in an architecture-specific sub-directory (e.g.
\file{amd64} on the x86 platform). If you wish to conform to this
convention for 64-bit JAGS then you should set the configure options
\option{--libdir}, \option{--libexecdir}, and \option{--bindir}
appropriately.

The configure script automatically detects the Sun Performance library,
which implements the BLAS/LAPACK functions.  

\clearpage
\section{Mac OS X}

There have been big changes in Apple's developer tools over the last few
years, but as of OS 10.9 things should hopefully be more stable.  These
instructions are for users that wish to build JAGS using clang and the
libc++ standard library, which have been the default since OS 10.9.  Users
needing to build against the libstdc++ library should see section \ref{section:otherosxoptions}, 
and users with an older version of OS X should refer to the installation
instructions given in older versions of the \JAGS~manual.

\subsection{Required tools}

If you wish to build from a released source package i.e.
\file{JAGS-\release.tar.gz}, you will need to install command line compilers
and agree to the license agreement before using them. The easiest way 
to do this is using the Terminal application from \file{/Applications/Utilities} 
- opening the application gives you a terminal with a UNIX shell known as bash.  
Run the command \code{xcode-select --install} on the terminal and follow the
instructions.  Then run the command \code{sudo xcodebuild -license} 
on the terminal, and follow the instructions to agree to the Apple
license terms (an administrators password is required).  You will then
need to install the gfortran package which you can download from:

\url{https://gcc.gnu.org/wiki/GFortranBinaries#MacOS}

This setup should be able to build the \JAGS\ sources and also source
packages in R.  All the necessary libraries such as BLAS and LAPACK are
included within OS X.  Additional tools are required to run the optional 
test suite (see section \ref{section:otherosxoptions}).


\subsection{Basic installation}


\subsubsection{Prepare the source code}

Move the downloaded \file{JAGS-\release.tar.gz} package to some suitable
working space on your disk and double click the file.  This will
decompress the package to give a folder called \file{JAGS-\release}.  
You now need to re-open the Terminal and
change the working directory to the \JAGS\ source code. In the Terminal
window after the \$ prompt type \code{cd} followed by a space.  In the Finder
drag \file{JAGS-\release} folder into the Terminal window and hit return.  If this
worked for you, typing \code{ls} followed by a return will list the contents
of the \JAGS\ folder.


\subsubsection{Set up the environment}
\label{section:osxenvironment}

The instructions below will create a configuration to build \JAGS\ using 
libc++ which will only run on 10.9 or later, and is therefore 64-bit only.
This is compatible with the ``Mavericks build'' of \R\ on CRAN and the 
corresponding \texttt{rjags} and \texttt{runjags} modules.  
In your Terminal window copy/paste the following code:

\begin{verbatim}
export CFLAGS="-Os -arch x86_64 -framework Accelerate"
export CXXFLAGS="-Os -arch x86_64 -framework Accelerate"
export FFLAGS="-Os -arch x86_64 -framework Accelerate"
export LDFLAGS="-arch x86_64"
\end{verbatim}

The \texttt{-framework Accelerate} finds the BLAS and LAPACK 
contained within the OS X installation.  The clang warning
\texttt{-framework Accelerate: `linker' input unused} can
be safely ignored.

\subsubsection{Configuration}

To configure the package type:

\begin{verbatim}
./configure
\end{verbatim}

This instruction should complete without reporting an error.

\subsubsection{Compile}
\label{section:osxcompile}

To compile the code type: 

\begin{verbatim} 
make -j 8 
\end{verbatim} 

The number `8' indicates the number of parallel build threads that
should be used (this will speed up the build process).  In general this
is best as twice the number of CPU cores in the computer - you may want
to change the number in the instruction to match your machine. Again,
this instruction should complete without errors.

\subsubsection{Install}
\label{section:osxinstall}

Finally to install \JAGS\ you need to be using an account with
administration privileges.  Type: 

\begin{verbatim}
sudo make install
\end{verbatim} 

This will ask for your account password and install the code ready to 
run as described in the User Manual. You need to ensure that
\texttt{/usr/local/bin} is in your PATH in order for the command 
\texttt{jags} to work from a shell prompt.

\subsection{Other installation options for Mac OS X}
\label{section:otherosxoptions}

\subsubsection{Running the test suite}

As of \JAGS\ version 4, a test suite is included with the source code that can be
run to ensure that the compiled code produces the expected results.  To run
this code on your installation, you will need to install the CppUnit framework - this can
be downloaded from \url{http://freedesktop.org/wiki/Software/cppunit/} or installed using
MacPorts or similar (see section \ref{section:osxtools}).  You should then be able to 
run the test suite from the Terminal  (in the \file{JAGS-\release} folder directory) by typing:

\begin{verbatim} 
make check
\end{verbatim} 

If successful, a summary of the checks will be given.  Note that the configuration and
first \texttt{make} step may need to be repeated if CppUnit was not installed the first
time these were run.  In this case, you may also need to clean the existing 
compiled code before running \texttt{make check} using:

\begin{verbatim} 
make clean
\end{verbatim} 


\subsubsection{Snow Leopard build}

The instructions below will
create a configuration to build a ``fat'' binary of \JAGS\ (containing
32-bit i386 and 64-bit x86\_64 code) using libstdc++ which will run on
OS 10.6 - 10.9.  This is compatible with the ``Snow Leopard build'' of
\R\ on CRAN and the corresponding \texttt{rjags} and \texttt{runjags} modules. 
In addition to the previously stated build requirements, installation of 
the (free) Xcode application from the Mac App store is required to be
able to follow these instructions.

In your Terminal window copy/paste the following code (note that there is no
space after 'platform' and the '$\backslash$' must not be followed by
any spaces): 

\begin{verbatim}
export CFLAGS="-Os -mmacosx-version-min=10.6 -isysroot \
/Applications/Xcode.app/Contents/Developer/Platforms/MacOSX.platform\
/Developer/SDKs/MacOSX10.9.sdk \
-stdlib=libstdc++ \
-arch i386 -arch x86_64"
export CXXFLAGS="-Os -mmacosx-version-min=10.6 -isysroot \
/Applications/Xcode.app/Contents/Developer/Platforms/MacOSX.platform\
/Developer/SDKs/MacOSX10.9.sdk \
-stdlib=libstdc++ \
-arch i386 -arch x86_64"
export FFLAGS="-Os -mmacosx-version-min=10.6 -isysroot \
/Applications/Xcode.app/Contents/Developer/Platforms/MacOSX.platform\
/Developer/SDKs/MacOSX10.9.sdk \
-arch i386 -arch x86_64"
export LDFLAGS="-mmacosx-version-min=10.6 -stdlib=libstdc++ \
-arch i386 -arch x86_64"
\end{verbatim}

Note that this configuration assumes you are running OS 10.9 - if you 
are on an earlier OS, you may need to modify the three occurances of 
\texttt{/Developer/SDKs/MacOSX10.9.sdk} to reflect your OS version.

You should then be able to follow the instructions given in sections \ref{section:osxcompile}
and \ref{section:osxinstall}.  If you have already used the same source directory to create
a Mavericks binary you will likely need to run \texttt{make clean} before switching
to build the Snow Leopard binary. If you encounter problems with these instructions on 
older versions of Mac OS X, consult previous versions of the installation manual 
which are contained within the disk images for older versions of \JAGS\ available from
\url{https://sourceforge.net/projects/mcmc-jags/files/JAGS/}


\subsection{Tips for developers and advanced users}

\subsubsection{Additional tools}
\label{section:osxtools}

Apple supplies a compiler suite, BLAS and LAPACK along with OS X developer tools, but some
additional tools are required to work with code from the \JAGS\ repository.  Probably 
the easiest way of obtaining the necessary utilities is using
\url{https://www.macports.org} or \url{http://brew.sh} 

The following sections assume that MacPorts is installed.

\subsubsection{Working with the development code}

If you want to work on code from the \JAGS\ repository, you will need to
build and install the auxillary GNU tools (autoconf, automake and
libtool), as well as mercurial, bison, flex and cppunit as follows:

\begin{verbatim}
sudo port install autoconf
sudo port install automake
sudo port install libtool
sudo port install mercurial
sudo port install bison
sudo port install flex
sudo port install cppunit
\end{verbatim}

Note that specific versions of some these ports may be required - the following 
are known to work:  autoconf 2.69, automake 1.14.1, libtool 2.4.6, 
mercurial 3.4, bison 3.0.4, flex 2.5.37 and cppunit 1.13.2.  Advice for
installing older ports (if necessary) is given at:

 \url{https://trac.macports.org/wiki/howto/InstallingOlderPort}

The following sequence should then retrieve a clone of the current
development branch of \JAGS, prepare the source code for
installation, compile the code and run the test suite:

\begin{verbatim} 
hg clone http://hg.code.sf.net/p/mcmc-jags/code-0 JAGS_devel
cd JAGS_devel
export CFLAGS="-Os -arch x86_64 -framework Accelerate"
export CXXFLAGS="-Os -arch x86_64 -framework Accelerate"
export FFLAGS="-Os -arch x86_64 -framework Accelerate"
export LDFLAGS="-arch x86_64"
autoreconf -fi
./configure
make -j 8
make check
\end{verbatim}

You should then be able to follow the install instructions given in section \ref{section:osxinstall}.

\subsubsection{Using ATLAS}

Rather than using the versions of BLAS and LAPACK provided within OS X, it
is possible to use ATLAS which can be installed via MacPorts using:

\begin{verbatim}
sudo port install atlas
\end{verbatim}

This command will take some hours to complete.  The \texttt{-framework Accelerate} flag can
then be omitted from the instructions given in section \ref{section:osxenvironment}.



\clearpage
\section{Windows}
\label{section:windows}

These instructions use MinGW, the Minimalist GNU system for Windows.
You need some familiarity with Unix in order to follow the build
instructions but, once built, \JAGS\ can be installed on any PC
running windows, where it can be run from the Windows command prompt.

\subsection{Preparing the build environment}

You need to install the following packages
\begin{itemize}
\item The Rtools compiler suite for Windows
\item MSYS  
\item NSIS, including the AccessControl plug-in  
\end{itemize}

\subsubsection{Rtools}

Rtools is a set of compilers and utilities used for compiling R on
Windows. Rtools can be downloaded from your nearest CRAN mirror
(\url{https://cran.r-project.org/bin/windows/Rtools/}).  We only need
the compilers, as we use the utilities provided by MSYS (See below).
For this reason, we choose not to add Rtools to the Windows environment
variable \code{PATH} when asked by the installer.

The JAGS binaries for Windows 4.0.0 and above are built with Rtools
3.3, which is based on gcc 4.6.3. We also successfully built JAGS with
the TDM-GCC compilers (\url{http://tdm-gcc.tdragon.net}) based on gcc
5.1.0. However, the resulting JAGS binary is not compatible with
R. The rjags package can be successfully compiled and linked against
JAGS built with TDM-GCC 5.1.0, and runs correctly on 64-bit R, but
the package spontaneously crashes in 32-bit R.

\subsubsection{MSYS}

MSYS (the Minimal SYStem) is part of the MinGW project (Minimal GNU
for Windows). It provides a bash shell for you to build Unix
software. Download the MinGW installer \file{mingw-get-setup.exe} from
\url{http://www.mingw.org}.  Run the installer and select
\code{msys-base} (``A Basic MSYS Installation (meta)'') for
installation and then select \code{Apply Changes} from the
\code{Installation} menu. There is no need to install the developer
toolkit (\code{mingw-developer-toolkit}) if you are working with a
release tarball of \JAGS.  You should not install any of the compilers
that come with MinGW as we shall be using the Rtools versions.

To make MSYS use the TDM compilers edit the file
\file{c:/mingw/msys/1.0/etc/fstab} to read
\begin{verbatim}
c:\Rtools\gcc-4.6.3\bin    /mingw
\end{verbatim}
This adds the Rtools compilers to your PATH inside the MSYS shell.

MSYS creates a home directory for you in
\file{c:/mingw/msys/1.0/home/username}, where \code{username} is your
user name under Windows. You will need to copy and paste the source
files for LAPACK and JAGS into this directory.

At the time of writing, the MinGW installer does not create a shortcut
for MSYS on either the desktop or the start menu, even when these
options are requested. Create your own shortcut to
\file{c:/MingGW/msys/1.0/msys.bat} which launches the MSYS shell. For
completeness, you may wish to use the icon
\file{c:/MinGW/msys/1.0/msys.ico} for your shortcut.

\subsubsection{NSIS}

The Nullsoft Scriptable Install System
(\url{http://nsis.sourceforge.net}) allows you to create a
self-extracting executable that installs \JAGS\ on the target PC.
These instructions were tested with NSIS 2.46.  You must also install
the AccessControl plug-in for NSIS, which is available from
\url{http://nsis.sourceforge.net/AccessControl_plug-in}. The plug-in
is distributed as a zip file which is unpacked into the installation
directory of NSIS.

\subsection{Building LAPACK}

Download the LAPACK source file from
\url{http://www.netlib.org/lapack} to your MSYS home directory. We
used version 3.5.0.

You need to build LAPACK twice: once for 32-bit \JAGS\ and once for
64-bit \JAGS. The instructions below are for 32-bit \JAGS. To build
64-bit versions, repeat the instructions with the flag \option{-m32}
replaced by \option{-m64} and start in a clean build directory.  Note
that you cannot cross-build 64-bit BLAS and LAPACK on a 32-bit Windows
system. This is because the build process must run some 64-bit test
programs.

Launch MSYS (\file{c:/MingW/msys/1.0/msys.bat}) and unpack the tarball.
\begin{verbatim}
tar xfvz lapack-3.5.0.tgz
cd lapack-3.5.0
\end{verbatim}
Copy the file \file{INSTALL/make.inc.gfortran} to \file{make.inc} in
the top level source directory.  Then edit \file{make.inc} replacing
the following lines:
\begin{verbatim}
FORTRAN = gfortran -m32
LOADER = gfortran -m32
\end{verbatim}
Type
\begin{verbatim}
make blaslib
make lapacklib 
\end{verbatim}
This will create two static libraries \file{librefblas.a} and
\file{liblapack.a}. These are insufficient for building \JAGS: you
need to create dynamic link library (DLL) for each one.

First create a definition file \file{libblas.def} that exports all the
symbols from the BLAS library
\begin{verbatim}
dlltool -z libblas.def --export-all-symbols librefblas.a
\end{verbatim}
Then link this with the static library to create a DLL
(\file{libblas.dll}) and an import library (\file{libblas.dll.a})
\begin{verbatim}
gcc -m32 -shared -o libblas.dll -Wl,--out-implib=libblas.dll.a \
libblas.def librefblas.a -lgfortran 
\end{verbatim}
 
Repeat the same steps for the LAPACK library, creating an import library
(\file{liblapack.dll.a}) and DLL (\file{liblapack.dll})
\begin{verbatim}
dlltool -z liblapack.def --export-all-symbols liblapack.a

gcc -m32 -shared -o liblapack.dll -Wl,--out-implib=liblapack.dll.a \
liblapack.def liblapack.a  -L./ -lblas -lgfortran 
\end{verbatim}

\subsection{Compiling \JAGS}

Unpack the JAGS source
\begin{verbatim}
tar xfvz JAGS-4.0.0.tar.gz
cd JAGS-4.0.0
\end{verbatim}
and configure JAGS for a 32-bit build
\begin{verbatim}
CC="gcc -m32" CXX="g++ -m32 -std=c++98" F77="gfortran -m32" \
LDFLAGS="-L/path/to/import/libs/ -Wl,--enable-auto-import" \
./configure
\end{verbatim}
where \file{/path/to/import/libs} is a directory that contains the
32-bit import libraries (\file{libblas.dll.a} and
\file{liblapack.dll.a}).  This must be an {\em absolute} path name,
and not relative to the JAGS build directory.


After the configure step, type
\begin{verbatim}
make win32-install
\end{verbatim}
This will install JAGS into the subdirectory \file{win/inst32}.
Note that you must go straight from the configure step to \texttt{make
  win32-install} without the usual step of typing \texttt{make} on
its own.  The \texttt{win32-install} target resets the installation
prefix, and this will cause an error if the source is already
compiled.

To install the 64-bit version, clean the build directory 
\begin{verbatim}
make clean 
\end{verbatim}
reconfigure JAGS for a 64-bit build:
\begin{verbatim}
CC="gcc -m64" CXX="g++ -m64" F77="gfortran -m64" \
LDFLAGS="-L/path/to/import/libs/ -Wl,--enable-auto-import" \
./configure
\end{verbatim}
Then type
\begin{verbatim}
make win64-install
\end{verbatim}
This will install JAGS into the subdirectory \file{win/inst64}.

With both 32-bit and 64-bit installations in place you can create the
installer.  Normally you will want to distribute the blas and lapack
libraries with JAGS.  In this case, put the 32-bit DLLs and import
libraries in the sub-directory \file{win/runtime32} and the 64-bit
DLLs and import libraries in the sub-directory
\file{win/runtime64}. They will be detected and included with the
distribution.

Make sure that the file \file{makensis.exe}, provided by
NSIS, is in your PATH. For a typical installation of NSIS, on 64-bit
windows:
\begin{verbatim}
PATH=$PATH:/c/Program\ Files\ \(x86\)/NSIS
\end{verbatim}
Then type
\begin{verbatim}
make installer
\end{verbatim}
After the build process finishes, the self extracting archive will be
in the subdirectory \file{win}.

\subsection{Running the unit tests}

In order to run the unit tests on Windows you must first install the
cppunit library from source. Download the file \file{cppunit-1.12.1.tar.gz}
from Sourceforge (\url{http://sourceforge.net/projects/cppunit/files/cppunit/1.12.1/}) and
unpack it:
\begin{verbatim}
tar xfvz cppunit-1.12.1.tar.gz
cd cppunit-1.12.1
\end{verbatim}
Then compile and install as follows:
\begin{verbatim}
CXX="g++ -m32" ./configure --prefix=/opt32 --disable-shared
make
make install
\end{verbatim}
The configure option \verb+--prefix=/opt32+ installs the 32-bit
library into \file{/opt32} instead of the default location
{/usr/local}. Using this option allows you to do a parallel
installation of the 64-bit version of the library, by rebuilding with
configure options \verb|CXX=g++ -m64| and \verb+--prefix=/opt64+. The two
installations will not interfere with each other.

The configure option \verb+--disable-shared+ prevents the creation of
the DLL \file{libccpunit.dll} and builds only the static library
\file{libcppunit.a}. Without this option, the unit tests will fail.
One of the major limitations of static linking to the C++ runtime is
that you cannot throw exceptions across a DLL boundary.  Linking the
test programs against \file{libcppunit.dll} will result in uncaught
exceptions and apparent failures for some of the tests, so it must be
disabled.\footnote{One of the attractions of the TDM-GCC compilers is
  that they do allow exceptions across DLL boundaries with static
  linking. However, we are not currently using TDM-GCC to build the
  JAGS binaries}.

To run the unit tests, add the option
\verb+--with-cppunit-prefix=/optXX+ when configuring JAGS where
\code{XX} is 32 or 64. Then run \code{make check} after \code{make
  winXX-install}.

\subsection{Running the examples}

The BUGS classic examples (file \file{classic-bugs.tar.gz} from the JAGS
Sourceforge site) can be run from the Windows command prompt using
the \command{make} command provided by Rtools. This requires adding Rtools
to the Windows search path if it is not currently there.
\begin{verbatim}
set PATH=c:\Rtools\bin;%PATH%
\end{verbatim}
You must have R installed, along with the packages \code{rjags} and
\code{coda}, both of which are available from CRAN
(\url{cran.r-project.org}).

It is necessary to add R to the search path and to set the variable
\verb+R_LIBS+. Note that here we are using the 64-bit version of R.
You may use the 32-bit version by substituting \code{i386} for
\code{x64}.
\begin{verbatim}
set PATH=c:\Program Files\R\R-3.2.2\bin\x64;%PATH%
set R_LIBS=c:\Users\username\Documents\R\win-library\3.2
\end{verbatim}
where \file{username} is your Windows user name. Then
\begin{verbatim}
tar xfvz classic-bugs.tar.gz
cd classic-bugs
cd vol1
make Rcheck
\end{verbatim}
will check all examples in volume 1 using the \code{rjags} package. Repeat
for \file{vol2} to complete the checks.

You can also run checks using the command line interface of JAGS. This
requires adding JAGS to the search path and overriding the default name
of the JAGS executable.
\begin{verbatim}
set PATH=c:\Program Files\JAGS\JAGS-4.0.0\bin\x64;%PATH%
set JAGS=jags.bat
\end{verbatim}
Then
\begin{verbatim}
make check
\end{verbatim}
in directory \file{vol1} or \file{vol2} will run the checks using the
command line interface.

\subsection{Using TDM-GCC compilers}

This section documents the use of TDM-GCC compilers to build JAGS.
TDM-GCC was used to build Windows binaries for JAGS 3.x.y, but has
been dropped in favour of Rtools for the 4.x.y release series. One
reason for this is that the 32-bit version of JAGS built with TDM-GCC
5.1.0 causes the rjags package to spontaneously crash. The 64-bit
version runs correctly.

TDM-GCC has a nice installer, available from Sourceforge (follow the
links on the main TDM-GCC web site
(\url{http://tdm-gcc.tdragon.net}). Ensure that you download the TDM64
MinGW-w64 edition as this is capable of producing both 32-bit and
64-bit binaries. We tested JAGS with \file{tdm64-gcc-5.1.0-2.exe}
based on gcc 5.1.0.

Select a ``Recommended C/C++'' installation and customize it by
selecting the Fortran compiler, which is not installed by default. The
installer gives you the option of adding TDM-GCC \file{bin} folder to
the windows PATH variable. We choose not to do this, but added the
\file{bin} to the PATH within the MSYS shell by editing
\file{c:/mingw/msys/1.0/etc/fstab} to read
\begin{verbatim}
c:\TDM-GCC-64    /mingw
\end{verbatim}
After installation of TDM-GCC, to force the compiler to use static
linking, delete any import libraries (files ending in \file{.dll.a} in
the TDM-GCC tree. If you do not do this then you will need to
distribute runtime DLLs from TDM-GCC with JAGS. You can easily do this
by copying the DLLs to \file{runtime32} and \file{runtime64} before
building the installer, as described above. Nevertheless, it is often
more convenient to use static linking.

Installation proceeds in the same way as for the Rtools build but with
two differences.  Firstly, when building the DLLs for blas and lapack,
you need to add the linker flag \verb+-lquadmath+ after
\verb+-lgfortran+. Secondly, when configuring JAGS you should set the
environment variable
\begin{verbatim}
CPPFLAGS=-D_GLIBCXX_USE_CXX11_ABI=0 
\end{verbatim}
This is necessary because gcc 5.1.0 introduced a new application
binary interface (ABI) for the C++ standard library (See
\url{https://gcc.gnu.org/onlinedocs/libstdc++/manual/using_dual_abi.html}. The
old ABI is still supported and is used if you set the above flag. If
you want to link JAGS with any software compiled with an earlier
version of gcc then you need to use the old ABI. Failure to do so will
result in error messages about undefined symbols from the linker.

\end{document}
